Robert C. Dunnell’s Systematics in Prehistory at 50

### 2021-06-07 analysis/paper/paper.Rmd
Text updates

## 2 SYSTEMATICS IN PALEOLITHIC ARCHAEOLOGY

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the ideosyncracies of archaeological classification are nowhere more apparent and acute than in Paleolithic archaeology. Rooted in French antiquarianism, the development of Palaeolithic systematics has been likened to 'accidents of history' @clark2009, and there are few periods or regions of the Palaeolithic that have not seen debate about the validity or otherwise of their analytical units @reynolds2019.

## 3 SYSTEMATICS IN PREHISTORY AND THE EMERGENCE OF EVOLUTIONARY ARCHAEOLOGY

For Dunnell himself, the soul-searching that began with writing *Systematics in Prehistory* led him to discover evolutionary theory. In a series of subsequent papers, he forcefully argued for the benefits of a scientific and Darwinian archaeology @dunnell1980; @dunnell1982. While not making many friends, these writings have since become foundational for the development of evolutionary archaeology. Initially, following the lead of Dunnell, this approach was rather narrowly selectionist - treating artefacts as the hard parts of the human phenotype and selection acting on these as the main driver of change @obrien1990 - but which since has become more fully aligned with cultural evolutionary thinking in the form of dual-inheritance and niche construction theory [@riede2019][@marwick2006].

>>>>

## 2 SYSTEMATICS IN PREHISTORY AND THE EMERGENCE OF EVOLUTIONARY ARCHAEOLOGY

For Dunnell himself, the soul-searching that began with writing *Systematics in Prehistory* led him to discover evolutionary theory. In a series of follow-up papers, he forcefully argued for the benefits of a scientific and Darwinian archaeology @dunnell1980; @dunnell1982. While not making him many friends, these writings have since become foundational for the development of evolutionary archaeology, especially in the Americas @obrien1996. Initially, following the diirect lead of Dunnell, this approach was rather narrowly selectionist - treating artefacts as the hard parts of the human phenotype and selection acting on these as the main driver of change @obrien1990 - but which since has become more fully aligned with cultural evolutionary thinking in the form of dual-inheritance and niche construction theory @riede2019 @marwick2006. Vitally, cultural evolutionary theory and its focus on the transmission of cultural knowledge via various modes of learning has provided the generative mechanism for material culture systematics. In its contemporary form, selection but also drift and a range of transmission biases play important roles in explaining culture change.

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the idiosyncrasies of archaeological classification are nowhere more apparent and acute than in the archaeology of human evolutionary history. Rooted in French antiquarianism, the development of Paleolithic systematics has been likened to 'accidents of history' @clark2009, and there are few periods or regions of the Paleolithic that have not seen debate about the validity or otherwise of their analytical units @reynolds2019. The use of, more often than not, older typological classifications remains prevalent, despite clear and repeated critiques (e.g. @bisson2000 ). More recently, the analysis of technological traits has supplemented or even eclipsed purely typological approaches yet theoretical explications of the generative mechanisms and rigorous comparative systematics backed by transparent and replicable analytics as demanded by Dunnell remain exceptions rather than the rule @tostevin2013. Today, a great deal attention is being paid to stematics in prehistory, and as Barton and Clark have pointed out, the continuing adherence to outmoded classifications is preventing the exploration of more relevant and pressing research questions @barton2021. Several researchers are tackling classificatory issues with novel and mostly quantitative means @ivanovait2020 @grove2021 @leplongeon2020. At the same time, however, it appears that the topic remains poorly heeded in archaeological teaching @lyman2021. Given that cultural evolutionary theory itself teaches us that aspects of culture most easily and rapidly change when scaffolded through active teaching, we suggest that, after half a century, Dunnell's *Systematics in Prehistory* - and with it rigorous and replicable ways of classifying material culture - are placed more abundantly on our curricula, in addition to continuing the ongoing critique and transformation of existing classifications through novel research. Only when the construction and meaning of our analytical units and their relationships among one another are transparent, and robust cultural taxonomies are in place can we seriously hope to understand the patterns and processes that have shaped cultural evolution in deep history.
